---
title: "607 week 7 assignment"
author: "Henry Owens"
date: "3/20/2021"
output:
  html_document: default
  pdf_document: default
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Working with HTML, XML, and JSON files in R

```{r, echo = FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(RCurl)
library(XML)
library(methods)
library(jsonlite)
library(knitr)
```

## Assigning the URLs

```{r}
url_html <- "https://raw.githubusercontent.com/hankowens/CUNY-MSDS/main/607/Week%207/Conan%20books%20-%20html.html" 
url_xml <- "https://raw.githubusercontent.com/hankowens/CUNY-MSDS/main/607/Week%207/conan%20books%20-%20xml.xml"
url_json <- "https://raw.githubusercontent.com/hankowens/CUNY-MSDS/main/607/Week%207/conan%20books%20-%20json.json"
```

## Importing the XML data

```{r xml}
get_xml <- xmlToDataFrame(getURL(url_xml))

get_xml
```

## Importing HTML data

```{r}
get_html <- getURL(url_html)
table_html <- as.data.frame(readHTMLTable(get_html, stringsAsFactors = FALSE))
table_html
```
## Importing JSON data

```{r}
raw_json <- fromJSON(getURL(url_json))

raw_json <- lapply(raw_json, function(x) {
  x[sapply(x, is.null)] <- NA
  unlist(x)})
table_json <- as.data.frame(do.call("rbind", raw_json))
table_json[1]
```

## Results

These three data formats result in similar but not identical tables in R. The way I did it, the XML was easiest with a single function to put into the data frame. HTML was a little more complicated, but JSON was the most complicated because it had an issue dealing with the null values where where was no Author2. 


```
