---
title: "Project 2 607 Owens Part 2"
author: "Henry Owens"
date: "3/15/2021"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(ggplot2)
library(gridExtra)
library(dplyr)
library(lubridate)
set.seed(2021)
options(digits=4)
```

## Tidying Data

I will look at the following data: 

* GDP Growth data from the world bank   
https://data.worldbank.org/indicator/NY.GDP.MKTP.KD.ZG?most_recent_year_desc=true

## World Bank GDP Growth Data
This data from the world bank describes annual gdp growth figures from just about every country and region from about 1940s to present although a lot countries didn't exist as independent countries in the 40s (Vietnam, Algeria, Ghana, etc.) and/or had no data until recently, so I am focusing on data since 2000. 

```{r cars, echo = TRUE}
url <- "https://raw.githubusercontent.com/hankowens/CUNY-MSDS/main/607/Tidy%20data/API_NY.GDP.MKTP.KD.ZG_DS2_en_csv_v2_2055665.csv"

worldbank <- read.csv(url, header = FALSE)
# Turn the row with years to char
worldbank[3,] <- as.character(worldbank[3,])
# drop first two useless rows
worldbank <- worldbank[-c(1,2),]
#rename header for country
worldbank[1,1] <- "country"
# eliminate years before 2000 (for simplicity)
worldbank <- worldbank[,-c(2:44)]
# set column headers and then delete first row
names(worldbank) <- as.matrix(worldbank[1, ])
worldbank <- worldbank[-1,]
# the last two rows are empty: deleting
worldbank <- worldbank[,-c(ncol(worldbank)-1, ncol(worldbank))]
```

## Pivoting the data to Tidy format

```{r}
wb_pivot <- worldbank %>% pivot_longer(!country, names_to = "year", values_to = "gdp_growth")
wb_pivot$gdp_growth <- as.numeric(wb_pivot$gdp_growth)
# Filtering NA Rows
wb_pivot <- wb_pivot %>% filter(!is.na(wb_pivot$gdp_growth)) 
```

```{r}
wb_groupby <- wb_pivot %>% 
  group_by(country) %>% 
  summarise(mean = mean(gdp_growth), sd = sd(gdp_growth), max = max(gdp_growth), 
            min = min(gdp_growth), n = n())
```

## Running hot

This is a very crude, naive investigation of GDP (because of compounding you shouldn't average growth rates, and I dont have the values to weight the rates, but I am curious if there is any relationship between having a high maximum GDP growth and average GDP growth. I wonder about the countries that have 25% or more GDP growth that is probably some kind of statistical anomaly; 125% certainly is. Not sure what to conclude about this but it is interesting. 


```{r}
ggplot(data = wb_groupby, mapping = aes(x = mean, y = max)) + 
  geom_point() +
  ggtitle("Mean and Maximum GDP Growth from 2000 to 2019")
```
